---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

<!-- badges: start -->
[![R-CMD-check](https://github.com/Olink-Proteomics/OlinkRPackage/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Olink-Proteomics/OlinkRPackage/actions/workflows/R-CMD-check.yaml)
[![lint](https://github.com/Olink-Proteomics/OlinkRPackage/actions/workflows/lint.yaml/badge.svg)](https://github.com/Olink-Proteomics/OlinkRPackage/actions/workflows/lint.yaml)
[![Coverage](https://img.shields.io/badge/Coverage-91%25-green)](https://img.shields.io/badge/Coverage-91%25-green)
[![CRAN status](https://www.r-pkg.org/badges/version/OlinkAnalyze)](https://CRAN.R-project.org/package=OlinkAnalyze)
[![](https://img.shields.io/badge/DOI-10.32614/CRAN.package.OlinkAnalyze-1f57b6?style=flat&link=https://doi.org/10.32614/CRAN.package.OlinkAnalyze)](https://doi.org/10.32614/CRAN.package.OlinkAnalyze)
<!-- badges: end -->

# Olink® Analyze

The goal of Olink® Analyze is to provide a versatile toolbox to enable easy and
smooth handling of Olink NPX data to speed up your proteomic research. Olink®
Analyze provides functions ranging from reading Olink NPX data as exported by
NPX Manager to various statistical tests and modelling, via different QC plot
functions. Thereby providing a convenient pipeline for your Olink NPX data
analysis.

## Installation

Olink® Analyze is available on CRAN:
https://cran.r-project.org/package=OlinkAnalyze.

```{r, eval = FALSE}
install.packages("OlinkAnalyze")
```

## Vignette

```{r, eval = FALSE}
browseVignettes("OlinkAnalyze")
```

**Note:** Process-specific, practical vignettes for analysis of Olink®
proteomics data hosted on OA have been moved to a newer package Olink® Analyze
Vignettes, that is hosted on CRAN:
https://cran.r-project.org/package=OlinkAnalyzeVignettes. Olink® Analyze
includes a master vignette and the core analysis functionality, and Olink®
Analyze Vignettes serves to complement this package.

```{r, eval = FALSE}
install.packages("OlinkAnalyzeVignettes")
```

## Usage

### Reading Olink NPX data

```{r read_data, eval = FALSE}
# open package
library(OlinkAnalyze)

# reading Olink NPX data
my_npx_data <- OlinkAnalyze::read_NPX(
  filename = "path/to/my_NPX_data.xlsx"
)
# OR
my_npx_data <- OlinkAnalyze::read_npx(
  filename = "path/to/my_NPX_data.xlsx"
)
```

### Check and clean NPX data

Olink® Analyze provides several functions to check and clean your NPX data.
Below follows an example of how to check and clean the NPX data using the
package provided `npx_data1` dataset:

```{r check_clean, eval = FALSE}
# check NPX data
check_npx_data1 <- OlinkAnalyze::check_npx(
  df = OlinkAnalyze::npx_data1
)

# clean NPX data
npx_data1_clean <- OlinkAnalyze::clean_npx(
  df = OlinkAnalyze::npx_data1,
  check_log = check_npx_data1
)

# re-check cleaned NPX data
check_npx_data1_clean <- OlinkAnalyze::check_npx(
  df = npx_data1_clean
)
```

### QC plot functions

There are several plot functions, below follow two examples using the package
provided `npx_data1` dataset:

```{r dist_plot, eval = FALSE}
# visualize the NPX distribution per sample per panel, example for one panel
npx_data1_clean |>
  dplyr::filter(
    .data[["Panel"]] == "Olink Cardiometabolic"
  ) |>
  OlinkAnalyze::olink_dist_plot(
    check_log = check_npx_data1_clean
  ) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_blank(),
    axis.ticks.x = ggplot2::element_blank()
  ) +
  ggplot2::scale_fill_manual(
    values = c("turquoise3", "red")
  )
```

```{r example_dist_plot, echo = FALSE}
knitr::include_graphics(
  path = normalizePath(
    path = "figures/example_distplot.png"
  ),
  error = FALSE
)
```

```{r qc_plot, eval = FALSE}
# visualize potential outliers by IQR vs. sample median per panel
# example for one panel
npx_data1_clean |>
  dplyr::filter(
    .data[["Panel"]] == "Olink Cardiometabolic"
  ) |>
  OlinkAnalyze::olink_qc_plot(
    check_log = check_npx_data1_clean
  ) +
  ggplot2::scale_color_manual(
    values = c("turquoise3", "red")
  )
```

```{r example_qc_plot, echo = FALSE}
knitr::include_graphics(
  path = normalizePath(
    path = "figures/example_qcplot.png"
  ),
  error = FALSE
)
```

### Normalization

Olink® Analyze provides several means of normalization when analyzing multiple
datasets. Below follows an example of reference sample (aka bridge)
normalization using the two package provided `npx_data1` and `npx_data2`
datasets:

```{r norm_example, eval = FALSE}
# identify bridge samples
bridge_samples <- intersect(
  x = npx_data1[["SampleID"]],
  y = npx_data2[["SampleID"]]
)
# remove control samples
bridge_samples <- bridge_samples[!grepl(
  pattern = "CONTROL",
  x = bridge_samples
)]

# npx_data1 was checked earlier
# we will check only npx_data2 before normalization

check_npx_data2 <- OlinkAnalyze::check_npx(
  df = npx_data2
)

# bridge normalize
bridge_normalized_data <- OlinkAnalyze::olink_normalization(
  df1 = npx_data1,
  df2 = npx_data2,
  overlapping_samples_df1 = bridge_samples,
  df1_project_nr = "20200001",
  df2_project_nr = "20200002",
  reference_project = "20200001",
  df1_check_log = check_npx_data1,
  df2_check_log = check_npx_data2
)
```

### Statistical tests and models

Olink® Analyze provides several statistical tests and model tools. Below follows
an example of how to perform a t-test and how to visualize the t-test output in
a volcano plot using the `npx_data1`:

```{r stats, eval = FALSE}
# t-test npx_data1
ttest_results_npx1 <- OlinkAnalyze::olink_ttest(
  df = npx_data1_clean,
  check_log = check_npx_data1_clean,
  variable = "Treatment"
)

# select names of the top #10 most significant proteins
ttest_sign_npx1 <- ttest_results_npx1 |>
  dplyr::slice_head(
    n = 10L
  ) |>
  dplyr::pull(
    .data[["OlinkID"]]
  )

# volcano plot with annotated top #10 most significant proteins
OlinkAnalyze::olink_volcano_plot(
  p.val_tbl = ttest_results_npx1,
  olinkid_list = ttest_sign_npx1
) +
  ggplot2::scale_color_manual(
    values = c("turquoise3", "red")
  )
```

```{r example_volcano_plot, echo = FALSE}
knitr::include_graphics(
  path = normalizePath(
    path = "figures/example_volcanoplot.png"
  ),
  error = FALSE
)
```

## Learn more

Please see the function specific help pages. Moreover, Olink® Analyze includes
two simulated NPX datasets for your convenience to help you explore the package
and its functions.

## Issues

Please report any issues (good or bad) to <biostattools[a]olink.com> or use the
github
[issue function](https://github.com/Olink-Proteomics/OlinkRPackage/issues).

## Alternative install methods

To install directly from the github repository:

```{r, eval = FALSE}
# Package remotes is required: install.packages("remotes")
remotes::install_github(
  repo = "Olink-Proteomics/OlinkRPackage/OlinkAnalyze",
  ref = "main",
  build_vignettes = TRUE
)
```

To install Olink Analyze into a new [conda](https://docs.conda.io/en/latest/)
environment:

```{bash, eval = FALSE}
conda create -n OlinkAnalyze -c conda-forge r-olinkanalyze
```

## Credits
Olink® Analyze is developed and maintained by the Olink Proteomics Data Science
Team.
