---
title: "Introduction to the Olink® Sample Index (OSI)"
output: 
  html_vignette:
    toc: true
    toc_depth: 3
    includes:
      in_header: ../man/figures/logo.html
vignette: >
  %\VignetteIndexEntry{Introduction to the Olink® Sample Index (OSI)}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  tidy = FALSE,
  tidy.opts = list(width.cutoff = 95),
  fig.width = 6,
  fig.height = 3,
  message = FALSE,
  warning = FALSE,
  time_it = TRUE,
  fig.align = "center"
)
```

Pre-analytical variation (PAV) refers to sample changes that occur prior
to measurement, for example due to variation in time from blood draw to
centrifugation or differences in temperature during sample handling.
These factors can alter the measured proteome without reflecting true
biology, and may therefore introduce unwanted variability into
proteomics studies.

The Olink® Sample Index (OSI) was developed to help users identify and
account for these pre-analytical effects in Olink® Explore HT EDTA
plasma datasets. OSI is a model-based framework derived from machine
learning models trained to recognize protein patterns associated with two
key handling variables: time to centrifugation and preparation
temperature. OSI values are included in the Olink® Explore HT NPX™ Map
output and are intended to support data quality assessment, exploratory
analysis, and downstream statistical modeling.


OSI provides multiple representations of the same underlying
pre-analytical signal. These include parameter-specific continuous
scores for time to centrifugation and preparation temperature, a
combined continuous summary score on a 0 to 1 scale, and a categorical
summary grouped into four ordinal levels. The summary variables are
useful for initial overview and communication, while the time and
temperature variables are more informative for deeper interpretation and
statistical correction.

This vignette introduces the OSI variables, explains how to interpret
them, and demonstrates how they can be used in exploratory plots and
statistical models. A practical workflow is to begin with the combined
OSI summary and OSI category, then investigate the individual time and
temperature components, and finally use OSITimeToCentrifugation and
OSITemperature as covariates in downstream analyses.

```{r, echo=FALSE}
library(OlinkAnalyze)
library(dplyr)
library(stringr)
library(ggplot2)
library(tidyr)
library(kableExtra)
```

```{r, echo=FALSE, include=F}
osi_data <- tibble(
  OSITimeToCentrifugation = c(
    0.3012289,0.060720572,0.94772694,0.720596273,0.142294296,
    0.549284656,0.954091239,0.585483353,0.404510282,0.647893479,
    0.319820617,0.307720011,0.219767631,0.369488866,0.984219203,
    0.154202301,0.091044,0.141906908,0.690007102,0.619256483,
    0.891394117,0.672999093,0.737077738,0.521135726,0.65983845,
    NA_real_,0.786281552,0.979821917,0.439431536,0.311702202,
    0.409474953,0.010467112,0.183849524,0.842729319,0.231161782,
    0.239099956,0.076691165,0.245723678,0.732135206,0.847453165,
    0.497527267,0.38790903,0.246448994,0.111096461,0.389994435,
    0.571935314,0.216892763,0.444768002,0.217990669,0.502299563,
    0.353904572,0.649985159,0.374713957,0.355445381,0.533687945,
    0.74033436,0.221102938,0.412746119,0.265686687,0.629973053,
    0.183828491,0.863644111,0.746568004,0.66828465,0.618017873,
    0.37223806,0.529835686,0.874682343,0.5817501,0.839767765,
    0.312448165,0.708290322,0.265017806,0.594343194,0.4812898,
    0.265032731,0.564590435,0.913188223,0.901874389,0.274166622,
    0.321482756,0.985640884,0.61999331,0.937314089,0.466532702,
    0.406832593,0.659230324,0.152346617,0.572867058,0.238726027,
    0.962358936,0.601365726,0.515029727,0.402573342,0.880246541,
    0.364091865,0.288239281,0.170645235,0.172171746,0.482042606
  ),
  
  OSIPreparationTemperature = c(
    0.252964929,0.21625479,0.674376388,0.047663627,0.700853087,
    0.351888638,0.408943998,0.820951324,0.918857348,0.28252833,
    0.961104794,0.728394428,0.686375082,0.052843943,0.395220135,
    0.47784538,0.560253264,0.698261595,0.915683538,0.618351227,
    0.428421509,0.542080367,0.058478489,0.260856857,0.397151953,
    0.197744737,0.831927563,0.152887223,0.803418542,0.546826157,
    0.662317642,0.171698494,0.63305536,0.311869747,0.724554346,
    0.398939825,0.969356411,0.967398371,0.726702539,0.257216746,
    NA_real_,0.593045652,0.267521432,0.531070399,0.785291671,
    0.168060811,0.404399181,0.471576278,0.868106807,0.925707956,
    0.881977559,0.674186843,0.950166979,0.516444894,0.576519021,
    0.336331206,0.347324631,0.020024301,0.502813046,0.871043414,
    0.006300784,0.072057124,0.164211225,0.770334074,0.735184306,
    0.971875636,0.466472377,0.074384513,0.648818124,0.75859317,
    0.137106081,0.396584595,0.224985329,0.057958561,0.395892688,
    0.0649283,0.225886433,0.054629109,0.67028204,0.297741783,
    0.100721582,0.071904097,0.880440569,0.754247402,0.816605888,
    0.982140374,0.103599645,0.099041829,0.798831611,0.784575267,
    0.009429905,0.779065883,0.729390652,0.630131853,0.48091083,
    0.156636851,0.00821552,0.452458394,0.492293329,0.389587112
  ),
  
  OSISummary = c(
    0.354783098,0.802812273,0.835708837,0.237749405,0.353986103,
    0.85688542,0.853763374,0.295895455,0.147048325,0.703992061,
    0.103806688,0.033727773,0.999404528,0.034874804,0.338391284,
    0.915063762,0.61723527,0.286285349,0.737797403,0.834054309,
    0.31427078,0.492566548,0.697373766,0.641462354,0.643922915,
    NA_real_,0.414735334,0.11940478,0.52602966,0.225073351,
    0.486411764,0.370214798,0.983350181,0.388319115,0.22924484,
    0.623297546,0.136540197,0.967469494,0.515071808,0.163070329,
    NA_real_,0.985954165,0.668771517,0.418915897,0.323344993,
    0.83525532,0.143817044,0.192815947,0.896738683,0.308119554,
    0.363300544,0.783946479,0.19337868,0.017765812,0.406607866,
    0.483167669,0.42184495,0.342808802,0.866483315,0.455108051,
    0.533764874,0.963843332,0.774591542,0.208876349,0.308786833,
    0.97134245,0.584900093,0.760823625,0.372709394,0.769193911,
    0.537677183,0.91399545,0.185296442,0.282218417,0.094962413,
    0.210487079,0.977098996,0.296302175,0.725983027,0.785687834,
    0.105417746,0.239594629,0.270544872,0.101058494,0.117913841,
    0.991236556,0.986054297,0.137067471,0.905309582,0.576301838,
    0.395448859,0.449802484,0.706501901,0.082502746,0.33931258,
    0.680787551,0.316949248,0.831568598,0.215172083,0.497948936
  ),
  
  OSICategory = c(
    2,4,4,1,2,4,4,2,1,3,
    1,1,4,1,2,4,3,2,3,4,
    2,2,3,3,3,0,2,1,3,1,
    2,2,4,2,1,3,1,4,3,1,
    0,4,3,2,2,4,1,1,4,2,
    2,4,1,1,2,2,2,2,4,2,
    3,4,4,1,2,4,3,4,2,4,
    3,4,1,2,1,1,4,2,3,4,
    1,1,2,1,1,4,4,1,4,3,
    2,2,3,1,2,3,2,4,1,2
  )
)

new_ids <- c(paste0("A", 1:77), paste0("B", 1:23))
stopifnot(nrow(osi_data) == length(new_ids))

osi_data <- osi_data |>
  mutate(SampleID = sample(new_ids, size = n(), replace = FALSE))

data1 <- OlinkAnalyze::npx_data1 |>
  dplyr::right_join(
    osi_data |>
      select(SampleID, OSITimeToCentrifugation, OSIPreparationTemperature,
             OSISummary, OSICategory),
    by = "SampleID"
  ) |>
  dplyr::filter(!grepl("CONTROL", SampleID))

temp_data <- olink_pca_plot(data1)

team <- temp_data[[1]]@data|> 
  arrange(PCX)|> 
  bind_cols(osi_data|> 
              arrange(OSITimeToCentrifugation)|> 
              select(-SampleID))

x <- rbeta(100, shape1 = 10, shape2 = 10) 

new_osi <- team|> 
  select(SampleID,OSITimeToCentrifugation,OSIPreparationTemperature,OSISummary,OSICategory)|> 
  mutate(OSIPreparationTemperature = x)|> 
  mutate(OSISummary = 0.6*OSITimeToCentrifugation + 0.4*OSIPreparationTemperature)|> 
  mutate(OSICategory = ifelse(
    is.na(OSISummary),
    0,
    cut(OSISummary,
        breaks = c(0, 0.25, 0.5, 0.75, 1),
        labels = 1:4,
        include.lowest = FALSE,  # ensures (0, 0.25]
        right = TRUE             # intervals are (a, b]
    )
  ))

npx_data <- OlinkAnalyze::npx_data1 |>
  dplyr::right_join(
    new_osi |>
      select(SampleID, OSITimeToCentrifugation, OSIPreparationTemperature,
             OSISummary, OSICategory),
    by = "SampleID"
  ) |>
  dplyr::filter(!grepl("CONTROL", SampleID))
```


## What is pre-analytical variation?

Pre-analytical variation arises when samples are handled differently
before measurement. In proteomics, this can occur because proteins and
other biomolecules continue to undergo biological and biochemical
processes after blood collection. Delays before centrifugation extend
the period during which cells and plasma remain in contact, while
temperature modulates the rate and nature of these processes. As a
result, samples from identical donors may show different protein
profiles solely because of differences in handling.

These effects are important because they may:

-   introduce systematic shifts in measured NPX values

-   create apparent outliers or clusters in exploratory analyses such as
PCA or UMAP

-   reduce power in downstream statistical analyses

-   confound biological comparisons if sample handling differs between
groups

PAV is therefore best viewed as a property of the sample and its
handling, rather than of the analytical platform itself. OSI is intended
to make these otherwise hidden handling-related patterns visible and
quantifiable.

## What is OSI?

The OSI is a model-based framework designed to
quantify key sources of pre-analytical variation in EDTA plasma samples
measured using Olink® Explore HT. The two modeled parameters are:

-   elapsed time between sample collection and centrifugation

-   preparation temperature prior to centrifugation

These were selected because they are broadly relevant, experimentally
validated, and consistently impactful sources of pre-analytical
variation in high-plex proteomics workflows.


OSI is not intended to reconstruct exact historical handling metadata.
Instead, it provides model-derived estimates that summarize how closely a
sample resembles handling-related patterns observed in controlled
studies. The aim is to support interpretation and robustness checks, not
to assign pass or fail status to samples.

## What OSI outputs are provided?

OSI provides several related outputs that can be used at different stages
of analysis. The recommended approach is to start with the combined
summary variables for an initial overview, then move to the
parameter-specific values for interpretation and statistical modeling.

```{r, echo=FALSE}
data.frame(
  Variable = c("OSISummary",
               "OSICategory",
               "OSITimeToCentrifugation",
               "OSITemperature"),
  Type = c("Continuous",
           "Categorical",
           "Continuous",
           "Continuous"),
  RecommendedUse = c("Initial QC overview and dataset-level visualization",
                     "Simple communication and quick stratification",
                     "Statistical correction and root-cause investigation",
                     "Statistical correction and root-cause investigation")
)|>
  kbl(booktabs = TRUE,
      caption = "Table 1. OSI variables")|>
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE,
                position = "center",
                latex_options = "HOLD_position")
```

### OSISummary

`OSISummary` is a continuous combined score that integrates information
from the time and temperature models into a single normalized measure on
a 0 to 1 scale. It is useful for initial QC, visualization, and
within-dataset comparison of overall pre-analytical impact. While lower
values are generally preferred, interpretation should focus on the
distribution and variability within the dataset rather than on absolute
cutoffs.

### OSICategory

`OSICategory` is a categorical version of the combined signal, grouped
into four ordinal levels from lower to higher impact. This simplified
representation is useful for communication and rapid assessment, but it
should not replace the continuous variables in detailed quantitative
analysis.

### OSITimeToCentrifugation

`OSITimeToCentrifugation` captures the time-related component of
pre-analytical variation. Lower values correspond more closely to
shorter delays before centrifugation. This variable is particularly
useful for understanding whether delayed sample processing may explain
variation in the data.

### OSITemperature

`OSITemperature` captures the temperature-related component of
pre-analytical variation. In contrast to the time variable, neither end
of the scale should be interpreted as universally better or worse on its
own. Instead, the key question is whether samples are handled
consistently or whether substructure in temperature-related patterns is
present within the dataset.

## Recommended OSI workflow

A practical workflow for OSI is:

-   Start with `OSISummary` and `OSICategory` to obtain a high-level
overview of pre-analytical variation in the dataset.

-   If variation, clustering, or outliers are observed, examine
`OSITimeToCentrifugation` and `OSITemperature` separately to
understand which handling dimension contributes to the observed
structure.

-   Use `OSITimeToCentrifugation` and `OSITemperature` as covariates in
downstream statistical models.

This reflects the intended use of OSI as a contextual QC and analysis
support tool rather than an automatic filter for excluding samples.

```{r, echo=FALSE}
npx_data <- read_NPX("~/NPX_file_location.xlsx")
```

For the remainder of this vignette, the examples assume that the data
have already been loaded into an object named npx_data.

### Initial overview with OSI summary variables

For an initial overview, it is helpful to inspect the PCA plot of the
`OSISummary` and the `OSICategory` values.

These plots provide a quick indication of whether pre-analytical
variation appears relatively homogeneous across the study or whether
subgroups and outliers may warrant closer investigation. OSI values are
interpreted relative to the dataset, so the overall pattern is usually
more informative than any single sample’s absolute value. The PCA plots could be created based on the `OSISummary` (Figure 1) and the `OSICategory` (Figure 2), separately.

```{r, echo=T,eval = T , message=FALSE, fig.cap= "Figure 1. PCA plot of the dataset colored by OSISummary."}
npx_data|>
  filter(!str_detect(SampleID, "CONTROL"))|>
  olink_pca_plot(color_g = "OSISummary")
```

```{r, echo=T,eval = T , message=FALSE, fig.cap= "Figure 2. PCA plot of the dataset colored by OSICategory."}
npx_data|>
  filter(!str_detect(SampleID, "CONTROL"))|>
  olink_pca_plot(color_g = "OSICategory")
```

From both PCA plots based on `OSISummary` and the `OSICategory`, we could see clear patterns. 

### Digging into OSI time and OSI temperature

After the summary variables have been reviewed, the next step is to
inspect the parameter-specific variables. These are often more useful
for identifying the likely source of the observed pre-analytical signal.

The same as OSI summary variables, PCA plots could be used to evaluate both `OSITimeToCentrifugation` (Figure 3) and `OSITemperature` (Figure 4).

```{r, echo=T,eval = T , message=FALSE, fig.cap= "Figure 3. PCA plot of the dataset colored by OSITimeToCentrifugation."}
npx_data|>
  filter(!str_detect(SampleID, "CONTROL"))|>
  olink_pca_plot(color_g = "OSITimeToCentrifugation")
```

```{r, echo=T,eval = T , message=FALSE, fig.cap= "Figure 4. PCA plot of the dataset colored by OSITemperature."}
npx_data|>
  filter(!str_detect(SampleID, "CONTROL"))|>
  olink_pca_plot(color_g = "OSITemperature")
```

Besides PCA plots, we could also visualize the OSI by density plots for both `OSITimeToCentrifugation` (Figure 5) and `OSITemperature` (Figure 6).

```{r, echo=T,eval = T , message=FALSE, fig.cap= "Figure 5. Density plot of the OSITimeToCentrifugation values."}
npx_data|>
  filter(!str_detect(SampleID, "CONTROL"))|>
  olink_osi_dist_plot(osi_score = "OSITimeToCentrifugation")
```

```{r, echo=T,eval = T , message=FALSE, fig.cap= "Figure 5. Density plot of the OSITemperature values."}
npx_data|>
  filter(!str_detect(SampleID, "CONTROL"))|>
  olink_osi_dist_plott(osi_score = "OSITemperature")
```

Compared to `OSITemperature`, `OSITimeToCentrifugation` displays greater heterogeneity 
and aligns with the structure observed in the PCA plot, indicating that it is 
likely the main contributor to the observed pre-analytical signal.

## How to use OSI in statistical models

For downstream statistical analysis, the recommended approach is to use
`OSITimeToCentrifugation` and `OSITemperature` as covariates. The
combined summary variables are more useful for initial QC and
visualization, whereas the individual time and temperature variables are
preferred for correction because they preserve more information about
the underlying handling dimensions. This is consistent with the
practical guidance in the OSI documentation. Since a clear pattern was 
identified from `OSITimeToCentrifugation`, taking ANOVA with `OSITimeToCentrifugation` 
as covariates for example:

```{r, echo=T,eval=F}
npx_data|>
  filter(SampleType == "SAMPLE")|>
  filter(AssayType == "assay")|> 
  olink_anova(variable = "Site",
              covariates = "OSITimeToCentrifugation")
```

A useful practice is to compare results with and without OSI
covariates. If conclusions remain stable, this supports the robustness
of the findings. If results change materially, then pre-analytical
variation may be contributing to the signal and should be reported and
interpreted transparently.


## How to interpret OSI values

Several principles are important when interpreting OSI:

-   OSI values are relative within a dataset and should not be compared
directly across datasets.

-   There is no universal threshold that defines good or poor OSI.

-   Uniform values across samples often indicate consistent handling,
even if the absolute values are not especially low.

-   Wide dispersion, substructure, or alignment with metadata such as
collection site or processing period may indicate that PAV should be
considered in downstream analyses.

-   OSITimeToCentrifugation is directional, where lower values
generally reflect shorter delays.

-   OSITemperature is not inherently directional in the same way, so
consistency is more informative than magnitude alone.

## Good to know and edge cases

The following practical considerations are important when using OSI:

-   OSI is currently validated for Olink® Explore HT.

-   OSI is scientifically validated for EDTA plasma and should not be
interpreted for other matrices in the same way.

-   OSI is intended for research use only and supports QC, study design,
and interpretation.

-   OSI should not be used as an automatic exclusion rule for samples or
cohorts.

-   OSI values are normalized within a dataset and are therefore not
directly comparable across studies.


## Limitations and considerations

OSI is highly useful, but it should be interpreted within its intended
scope.

-   It is limited to the modeled dimensions of time to centrifugation
and preparation temperature.

-   It does not replace true handling metadata.

-   It should be interpreted with caution when samples are suspected to
fall outside the conditions represented in model development.

-   It is not intended to retroactively invalidate previously completed
studies.

-   Associations between OSI and biology may still arise if biological
groups were handled differently during sample collection or
processing.


Summary

Pre-analytical variation is a common and important source of variability
in proteomics studies. The OSI provides a
structured way to detect and account for handling-related effects in
Olink® Explore HT EDTA plasma datasets. OSI is most effective when used
as part of a workflow that begins with OSISummary and OSICategory
for initial QC, then moves to OSITimeToCentrifugation and
OSITemperature for interpretation and statistical adjustment.

In practice, the most important question is not whether a sample has a
high or low absolute OSI value, but whether OSI reveals heterogeneity,
substructure, or confounding that could influence interpretation of the
data. Used in this way, OSI can improve transparency, support robust
analysis, and increase confidence in biological conclusions.


## Contact Us

We are always happy to help. Email us with any questions:

-   biostat\@olink.com for statistical services and general stats
    questions

-   support\@olink.com for Olink lab product and technical support

-   info\@olink.com for more information

## Legal Disclaimer

© 2025 Olink Proteomics AB, part of Thermo Fisher Scientific.

Olink products and services are For Research Use Only. Not for use in
diagnostic procedures.

All information in this document is subject to change without notice.
This document is not intended to convey any warranties, representations
and/or recommendations of any kind, unless such warranties,
representations and/or recommendations are explicitly stated.

Olink assumes no liability arising from a prospective reader’s actions
based on this document.

OLINK, NPX, PEA, PROXIMITY EXTENSION, INSIGHT and the Olink logotype are
trademarks registered, or pending registration, by Olink Proteomics AB.
All third-party trademarks are the property of their respective owners.

Olink products and assay methods are covered by several patents and
patent applications
[https://www.olink.com/patents/](https://olink.com/patents/).
