---
title: "Introduction to bridging Olink® NPX datasets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to bridging Olink® NPX datasets}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  tidy = FALSE,
  tidy.opts = list(width.cutoff = 95),
  fig.width = 10,
  message = FALSE,
  warning = FALSE,
  time_it = TRUE,
  fig.width = 7
)
```

Individual Olink® NPX datasets are normalized using either plate control
normalization or intensity normalization methods. Plate control
normalization is generally used for single plate projects, while
intensity normalization is used for multiplate projects. Additionally,
intensity normalization method assumes that all samples within a project
are fully randomized.

In the case where all samples within a project are not fully randomized,
or when a study is separated into separate batches, an additional
normalization step is needed to allow the data to be comparable, since
NPX is a relative measurement. The joint analysis of two or more Olink®
NPX datasets often requires an additional batch correction step to
remove technical variations, which is referred to as bridging.

Bridging is also needed if Olink® NPX datasets are:

-   plate control normalized only and run conditions (e.g lab and
    reagent lots) have changes.

-   intensity normalized but from two different sample populations.

To bridge two or more Olink® NPX datasets, bridging samples are needed
to calculate the assay-specific adjustment factors between datasets.
Bridging samples are shared samples among datasets - that is that
samples that are analysed in both datasets. The recommended number of
bridging samples are shown in the table below. Olink® NPX datasets
without shared samples should not be combined using the bridging
approach described below.

```{r, echo=FALSE}
library(OlinkAnalyze)
library(dplyr)
library(stringr)
library(ggplot2)
library(kableExtra)
```

```{r brnrtab, message=FALSE, echo=FALSE}
data.frame(Platform = c("Target 96",
                         "Explore 1536",
                        "Explore Expansion"),
           BridgingSamples = c("8-16",
                               "8-16",
                              "16-24")) |>
  kbl(booktabs = TRUE,
      digits = 2,
      caption = "Recommended number of bridging samples for Olink platforms") |>
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE,
                position = "center",
                latex_options = "HOLD_position")

```

The following tutorial is designed to give you an overview of the kinds
of data combining methods that are possible using the Olink® bridging
procedure. Before starting bridging, it is important to check if the
same sample IDs were assigned to the bridging samples.

## Selecting bridging samples

Prior to running the second study, bridging samples must be selected
from the reference study and be added to the second study. These samples
can be selected using the `olink_bridgeselector()` function in Olink
Analyze. The bridge selection function will select a number of bridge
samples based on the reference data. This function selects samples
which passes QC and have high detectibility. To select samples across
the range of the data, the samples are ordered by mean NPX value and
selected across this range. When running the selector, Olink recommends
starting at sampleMissingFreq = 0.10 which represents a maximum of 10%
data below LOD per sample. If there are not enough samples output,
increase to 20%. For alternative matrices and specific disease types, it
may be needed to increase the sampleMissingFreq to higher levels.

In this example we will demonstrate how to select 16 bridging samples
using `npx_data1` which will act as the reference data.

```{r bridge_sample_selection}
olink_bridgeselector(df = npx_data1,
                     sampleMissingFreq = 0.1,
                     n = 16) |>
  kableExtra::kbl(booktabs = TRUE,
      digits = 2,
      caption = "Selected Bridging Samples") |>
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE,
                position = "center", latex_options = "HOLD_position")
  
```

## Setup bridging datasets

Bridging datasets are standard Olink® NPX tables. They can be loaded
using `read_NPX()` function with default Olink Software NPX file as
input.

```{r message=FALSE, eval=FALSE, echo = TRUE}
data1 <- read_NPX("~/NPX_file1_location.xlsx")
data2 <- read_NPX("~/NPX_file2_location.xlsx")
```

To demonstrate how bridging works, we will use the example datasets
(`npx_data1` and `npx_data2`) from **Olink Analyze** package. This
workflow also uses functions from the dplyr, stringr, and ggplot2
packages.

## Check bridging datasets

First, confirm that there are overlapping sample IDs within the study.
It is important that the sample IDs are the same in both NPX files.

```{r}

data.frame(SampleID = intersect(npx_data1$SampleID, npx_data2$SampleID)) |>
  dplyr::filter(!stringr::str_detect(SampleID, "CONTROL_SAMPLE")) |> #Remove control samples
  kableExtra::kbl(booktabs = TRUE,
      digits = 2,
      caption = "Overlapping Bridging Samples") |>
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE,
                position = "center", latex_options = "HOLD_position")
```

Then, gain an overview of the datasets that are going to be bridged. For
example, plot and compare NPX distribution between datasets. By having a
sense of how the studies compared to each other before bridging, we can
then determine the success of the bridging process post bridging.

```{r check, message=FALSE}
# Load datasets
npx_1 <- npx_data1 |>
  filter(!str_detect(SampleID, "CONTROL_SAMPLE")) |> #Remove control samples
  mutate(dataset = "data1")
npx_2 <- npx_data2 |>
  filter(!str_detect(SampleID, "CONTROL_SAMPLE")) |> #Remove control samples
  mutate(dataset = "data2")

npx_df <- bind_rows(npx_1, npx_2)


# Plot NPX density before bridging normalization
npx_df |>
  mutate(Panel = gsub("Olink ", "", Panel)) |>
  ggplot(aes(x = NPX, fill = dataset)) +
  geom_density(alpha = 0.4) +
  facet_grid(~Panel) +
  olink_fill_discrete(coloroption = c("red", "darkblue")) +
  set_plot_theme() +
  ggtitle("Before bridging normalization: NPX distribution") +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        strip.text = element_text(size = 16),
        legend.title = element_blank(),
        legend.position = "top")
```

Use a PCA plot to visualize sample-to-sample distance before bridging.
Typically the project dataset accounts for most of the observed
variation within the combined datasets at this point.

```{r pca1, message=FALSE, fig.cap="PCA plot of combined datasets before bridging"}
## before bridging

#### Extract bridging samples
npx_1 <- npx_data1 |>
  mutate(dataset = "data1")
npx_2 <- npx_data2 |>
  mutate(dataset = "data2")

overlap_samples <-  data.frame(SampleID = intersect(npx_1$SampleID, npx_2$SampleID)) |>  
  filter(!str_detect(SampleID, "CONTROL_SAMPLE")) |> #Remove control samples
  pull(SampleID)


### Generate unique SampleIDs
npx_1 <- npx_data1 |>
  mutate(dataset  = "data1",
         SampleID = paste0(dataset, PlateID, SampleID))

br_1 <- npx_data1 |>
  mutate(dataset = "data1") |>
  filter(SampleID %in% overlap_samples) |>
  mutate(SampleID = paste0(dataset, PlateID, SampleID))

npx_2 <- npx_data2 |>
  mutate(dataset  = "data2",
         SampleID = paste0(dataset, PlateID, SampleID))

br_2 <- npx_data2 |>
  mutate(dataset = "data2") |>
  filter(SampleID %in% overlap_samples) |>
  mutate(SampleID = paste0(dataset, PlateID, SampleID))

npx_before_br <- bind_rows(npx_1, npx_2) |>
  mutate(datatype = ifelse(SampleID %in% c(br_1$SampleID, br_2$SampleID),
                           "bridges", dataset))



### PCA plot
OlinkAnalyze::olink_pca_plot(df          = npx_before_br,
                             color_g     = "datatype",
                             byPanel     = TRUE)

```

## Perform bridging between two data sets

We can use `olink_normalization()` function to bridge two datasets. The
bridging procedure is to first calculate **the median of the paired NPX
differences** per assay between the bridging samples as adjustment
factor then use these adjustment factors to adjust NPX values between
two datasets. In this process, one dataset is considered the reference
dataset (`df1`) and its NPX values remain unaltered. The other dataset
is considered the new dataset (`df2`) and is adjusted to the reference
dataset based on the adjustment factors.

The output from `olink_normalization()` function is a NPX table with
adjusted NPX value in the column `NPX`.

`olink_normalization()` uses the information from column `Project` to
distinguish between reference dataset from the other dataset. It is up
to the user to define which dataset is the reference dataset and specify
the names of the bridge samples. The resulting dataset will contain the
reference dataset, which will be identical to the input reference data,
with adjustment factors of 0, and the newly bridged dataset.

```{r bridging, message=FALSE}
# Find shared samples
npx_1 <- npx_data1 |>
  mutate(dataset = "data1")
npx_2 <- npx_data2 |>
  mutate(dataset = "data2")

overlap_samples <-data.frame(SampleID = intersect(npx_1$SampleID, npx_2$SampleID)) |>
  filter(!str_detect(SampleID, "CONTROL_SAMPLE")) |> #Remove control samples
  pull(SampleID)

# Perform Bridging normalization
npx_br_data <- olink_normalization(df1                     = npx_1,
                                   df2                     = npx_2,
                                   overlapping_samples_df1 = overlap_samples,
                                   df1_project_nr          = "20200001",
                                   df2_project_nr          = "20200002",
                                   reference_project       = "20200001")
dplyr::glimpse(npx_br_data)

```

## Perform evaluation of bridging normalization

First, check NPX distribution in datasets after bridging normalization.

```{r densitybr, message=FALSE}
# Plot NPX density after bridging normalization

npx_br_data |>
  mutate(Panel = gsub("Olink ", "", Panel)) |>
  ggplot2::ggplot(ggplot2::aes(x = NPX, fill = dataset)) +
  ggplot2::geom_density(alpha = 0.4) +
  ggplot2::facet_grid(~Panel) +
  olink_fill_discrete(coloroption = c("red", "darkblue")) +
  set_plot_theme() +
  ggplot2::ggtitle("After bridging normalization: NPX distribution") +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        strip.text = element_text(size = 16),
        legend.title = element_blank(),
        legend.position = "top")

```

Then summarize number of assays that have adjustment factors in certain
ranges. High adjustment factors can result from variations between
projects, such as panel versions or technical modifications. Such assays
can be visualized individually with violin plots and may warrant further
investigation to confirm they are still comparable between projects.

```{r tbl_adj_fct, message=FALSE}
breaks <- c(-Inf, -1, -0.25, 0.25, 1, Inf)
labels <- c("[< -1]", "(-1,-0.25]", "(-0.25,0.25]", "(0.25,1]", "(>1]")

# reference project defined in bridging normalization
reference_project <- "20200001"
npx_br_data |>
  mutate(Panel = gsub("Olink ", "", Panel)) |>
  # remove adjustment factors from reference project.
  filter(Project != reference_project) |>
  mutate(group = cut(Adj_factor, breaks = breaks,
            labels = labels)) |>
  select(OlinkID, Adj_factor, Panel, group) |>
  distinct() |>
  group_by(Panel, group) |>
  tally() |>
  tidyr::pivot_wider(names_from = group, values_from = n) |>
  tibble::column_to_rownames(var = "Panel") |>
  dplyr::relocate(any_of(labels)) |>
  kableExtra::kbl(booktabs = TRUE,
      digits = 2,
      caption = paste("Distribution of assays in different ranges of",
                      "adjustment factors")) |>
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE,
                position = "center", latex_options = "HOLD_position")

```

```{r voilin plot}
# High Adjustment Factor Assays
cutoff <- 1

high_assay<-npx_br_data  |> 
 filter(Adj_factor > cutoff | Adj_factor < -cutoff) |> 
  select(Assay, OlinkID, Panel) |>
  distinct()
npx_data1 |>
  mutate(Project = "20200001") |>
  rbind({
    npx_data2 |>
      mutate(Project = "20200002")
  }) |>
  filter(OlinkID %in% high_assay$OlinkID[1]) |>
  filter(!str_detect(SampleID, "CONTROL*.")) |>
  mutate(Assay_OID = paste(Assay, OlinkID, sep = "\n")) |> 
  ggplot2::ggplot(aes(Project, NPX)) +
  ggplot2::geom_violin(aes(fill = Project)) +
  theme(legend.position = "none") +
  set_plot_theme() +
  facet_wrap(. ~ Assay_OID, scales='free_y')
```

Finally, use PCA plot to check whether bridging normalization has effect
in correcting batch effects. In the example below, it is clear that
before bridging samples from data 1 and 2 are divided into separate
clusters due to the batch effects, but after bridging they are shown as
one cluster in the PCA plot. Bridging normalization has sufficiently
removed the batch effects between two data sets.

```{r pca2, message=FALSE, fig.cap="PCA plot of combined datasets after bridging"}
## After bridging

### Generate unique SampleIDs
br_ids <- npx_before_br |>
  select(SampleID, OlinkID, datatype)
npx_after_br <- npx_br_data |>
  mutate(SampleID = paste0(dataset, PlateID, SampleID)) |>
  left_join(br_ids, by = c("SampleID", "OlinkID"))
### PCA plot
OlinkAnalyze::olink_pca_plot(df          = npx_after_br,
                             color_g     = "datatype",
                             byPanel     = TRUE)
```

## Contact us

We are always happy to help. Email us with any questions:

biostat\@olink.com for statistical services and general stats questions

biostattools\@olink.com for Olink Analyze and Shiny app support

support\@olink.com for Olink lab product and technical support

info\@olink.com for more information
