---
title: "Bridging Olink^®^ Explore 3072 to Olink^®^ Explore HT"
output: 
  html_vignette:
    toc: true
    toc_depth: 3
    includes:
      in_header: ../man/figures/logo.html
vignette: >
  %\VignetteIndexEntry{Bridging Olink^®^ Explore 3072 to Olink^®^ Explore HT}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  tidy = FALSE,
  tidy.opts = list(width.cutoff = 95),
  fig.width = 6,
  fig.height = 3,
  message = FALSE,
  warning = FALSE,
  time_it = TRUE,
  fig.align = "center"
)
```

```{r, echo=FALSE}
library(OlinkAnalyze)
library(dplyr)
library(stringr)
library(ggplot2)
library(kableExtra)
```

## Introduction

## Bridge Sample Selection

## Import NPX files

The input datasets are standard Olink^®^ NPX^TM^ tables. They can be loaded
using `read_NPX()` function with default Olink Software NPX file as
input.

```{r message=FALSE, eval=FALSE, echo = TRUE}
data_E3072 <- read_NPX("~/NPX_Explore3072_location.parquet") # Could also be a CSV file
data_EHT <- read_NPX("~/NPX_ExploreHT_location.parquet")
```

## Checking input datasets and bridging samples

## Normalization

### Is the Assay Bridgeable

### Normalization using the median of paired differences

If it is expected that both the kind of distribution and the equal variance per test between runs are the same, then the normalization using the median of paired differences will be preferred. The normalization using the median of paired differences based on the bridge samples is performed in the following steps:

  1. For each assay and the Explore 3072 project, calculate the pairwise difference for each of the overlapping samples with the Explore HT project.
  
  2. Estimate the normalization factor of each assay by finding the median of the pairwise differences.
  
  3: Use the assay-specific normalization factor for each assay to normalize each data point from Explore 3072 to Explore HT.

### Quantile-Quantile normalization

Since Explore HT and Explore 3072 are two distinct products and the assays differ considerably between products, some of the assays exist in corresponding but distinct NPX spaces. For those assays, a Quantile-Quantile normalization will be favored. The normalization using the Quantile-Quantile Normalization based on the bridge samples is performed in the following steps:

  1. For each assay of Explore 3072, calculate the empirical distribution function.
  
  2. Map the empirical distribution function from Explore 3072 to the Explore HT space using the quantiles.
  
  3. Construct a spline regression model using the sorted Explore 3072 data and the mapped Explore 3072 data along with the anchor points of the spline function.
  
  4. Use the spline regression model to predict all the data points from Explore 3072 to Explore HT.


### Function Output

## Evaluating the quality of bridging

## Export Normalized Data

## FAQs

### Correlation Assays

### Downstream Analysis


## Contact Us

We are always happy to help. Email us with any questions:

-   biostat\@olink.com for statistical services and general stats
    questions

-   support\@olink.com for Olink lab product and technical support

-   info\@olink.com for more information

## Legal Disclaimer

© 2024 Olink Proteomics AB.

Olink products and services are For Research Use Only and not for Use in
Diagnostic Procedures.

All information in this document is subject to change without notice.
This document is not intended to convey any warranties, representations
and/or recommendations of any kind, unless such warranties,
representations and/or recommendations are explicitly stated.

Olink assumes no liability arising from a prospective reader’s actions
based on this document.

OLINK, NPX, PEA, PROXIMITY EXTENSION, INSIGHT and the Olink logotype are
trademarks registered, or pending registration, by Olink Proteomics AB.
All third-party trademarks are the property of their respective owners.

Olink products and assay methods are covered by several patents and
patent applications <https://olink.com/legal/patents>.

