---
title: "Introduction to bridging datasets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to bridging datasets}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  fig.width = 10,
  message = FALSE,
  warning = FALSE,
  time_it = TRUE,
  fig.width = 7
)
```

# Introduction to bridging Olink®NPX datasets

Olink® NPX datasets are normalized datasets using either plate control
normalization or intensity normalization methods. Intensity
normalization method assumes that all samples within a project are fully
randomized.

The joint analysis of two or more Olink® NPX datasets often requires
additional batch correction step to remove technical variations, which
is referred to as bridging.

Bridging is needed if Olink® NPX datasets are:

-   plate control normalized only and run conditions (e.g lab and
    reagent lots) have changes.

-   intensity normalized but from two different sample populations.

To bridge two or more Olink® NPX datasets, bridging samples are needed
to calculate adjustment factors between datasets. The recommended number
of bridging samples for Explore 1536 datasets is between 8-16.

The following tutorial is designed to give you an overview of the kinds
of data combining methods that are possible using the Olink® bridging
procedure.

## Setup the bridging objects

To demonstrate how bridging works, we will use the example datasets
(`npx_data1` and `npx_data2`) from **OlinkAnalyze** package. They are in
standard Olink® NPX table format that Olink® delivers to customers.

```{r setup}
library(OlinkAnalyze)
library(dplyr)
library(stringr)
```

## Perform bridging

We can use `olink_normalization()` function to bridge two datasets. The bridging 
procedure is to first calculate **the median of the paired NPX differences** per
assay between the bridging samples as adjustment factor then use these adjustment 
factors to adjust NPX values between two datasets. The output from 
`olink_normalization()` function is a table with adjusted NPX value in the column 
`NPX`.

```{r bridging, message=FALSE}
           
# Find overlapping samples
npx_1 <- npx_data1 %>% 
  mutate(dataset = "data1")
npx_2 <- npx_data2 %>% 
  mutate(dataset = "data2")

overlap_samples <- intersect(npx_1$SampleID, npx_2$SampleID) %>% 
  data.frame() %>% 
  filter(!str_detect(., 'CONTROL_SAMPLE')) %>% #Remove control samples
  pull(.)
# Perform Bridging normalization
npx_br_data <- olink_normalization(df1 = npx_1, 
                                   df2 = npx_2, 
                                   overlapping_samples_df1 = overlap_samples,
                                   df1_project_nr = '20200001',
                                   df2_project_nr = '20200002',
                                   reference_project = '20200001')
glimpse(npx_br_data)

```

## Perform evaluation of bridging normalization

PCA plot is used to visualize sample-to-sample distance before and after bridging.

```{r pca1, message=FALSE, fig.cap="PCA plot of combined datasets without bridging"}
## before bridging
npx_1 <- npx_data1 %>% 
  mutate(dataset = "data1") %>% 
  mutate(SampleID = paste0(dataset, PlateID, SampleID))
  
npx_2 <- npx_data2 %>% 
  mutate(dataset = "data2") %>% 
  mutate(SampleID = paste0(dataset, PlateID, SampleID))
             
npx_data <- rbind(npx_1, npx_2)
OlinkAnalyze::olink_pca_plot(df = npx_data,
                               color_g = "dataset", byPanel = TRUE, coloroption = c('orange', 'darkblue')) 

```


```{r pca2, message=FALSE, fig.cap="PCA plot of combined datasets after bridging"}
## After bridging
npx_br_plot <- npx_br_data %>% 
  mutate(SampleID = paste0(dataset, PlateID, SampleID))
OlinkAnalyze::olink_pca_plot(df = npx_br_plot,
                               color_g = "dataset", byPanel = TRUE, coloroption = c('orange', 'darkblue')) 
```

