---
title: "Bridging Olink^®^ Explore 3072 to Olink^®^ Explore HT"
output: 
  html_vignette:
    toc: true
    toc_depth: 3
    includes:
      in_header: ../man/figures/logo.html
vignette: >
  %\VignetteIndexEntry{Bridging Olink^®^ Explore 3072 to Olink^®^ Explore HT}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  tidy = FALSE,
  tidy.opts = list(width.cutoff = 95),
  fig.width = 6,
  fig.height = 3,
  message = FALSE,
  warning = FALSE,
  time_it = TRUE,
  fig.align = "center"
)
```

```{r, echo=FALSE}
library(OlinkAnalyze)
library(dplyr)
library(stringr)
library(ggplot2)
library(kableExtra)
```

## Introduction

Individual Olink® NPX datasets are normalized using either plate control
normalization or intensity normalization methods. In the case when a study is 
separated into multiple batches, an additional normalization step is needed to 
allow the data to be comparable, since NPX is a relative measurement. The 
following tutorial is designed to give you an overview of the Olink® bridging 
procedure for combining datasets from Explore 3072 and Explore HT products.

### Within- and between-product bridging

The joint analysis of two or more Olink® NPX datasets run on the same Olink® 
product often requires an additional batch correction step to
remove technical variation, which is referred to as overlapping sample 
reference normalization, bridge normalization, or just simply bridging. For more
information on within-product bridging, see the [Introduction to Bridging 
tutorial](https://cran.r-project.org/web/packages/OlinkAnalyze/vignettes/bridging_introduction.html).


Explore 3072 and Explore HT are both Olink® platforms that use PEA technology
combined with next generation sequencing to calculate NPX for thousands of 
proteins. Since many of the assays profiled in Explore 3072 are also found on 
Explore HT, bridging data across products enables increased power in studies 
consisting of both 3072 and HT datasets, rather than limiting these studies to 
meta-analysis. However, these products differ in the number of assays measured 
and in the **antibodies and reagents used**, which means that normalization 
cannot be performed in the same way that within-product bridging is performed.

In the case where a study consists of separate batches run in both the Olink® 
Explore 3072 and the Explore HT platforms, an additional batch correction
step is required to allow data from these two platforms to be analyzed together, 
which is referred to as **product bridging, or Explore 3072 to HT bridging**.

### Considerations for product bridging

Product bridging allows the NPX values of an Olink® Explore 3072 project to be 
normalized and made comparable to the NPX values of an Olink® Explore HT project. This
process is one-directional, and normalizing Explore HT NPX values to Explore 
3072 is not supported. 

The product bridging normalization uses the **~3000** assays that are
overlapping between Explore 3072 and HT. Each overlapping assay undergoes 
a series of checks that evaluate the number of counts, correlation, and 
difference of NPX ranges between the two datasets. If an assay has enough counts
and comparable metrics between the two datasets, it is determined to be suitable
for bridging (referred to as a "bridgeable" assay). Depending on the NPX 
distribution of each bridgeable assay in the two datasets, the assay is
normalized using either median normalization or **QQ normalization**.

Product bridging between an Explore 3072 and an Explore HT NPX dataset
requires **40-60 bridging samples**. Bridging samples are shared samples among
datasets - that is that samples that are analyzed in both datasets. Olink® NPX
datasets without shared samples should not be combined using the bridging 
approach described below. Before starting bridging, it is important to check if 
the same sample IDs were assigned to the bridging samples. More information on 
bridge sample selection can be found in the [selecting bridging samples section](https://cran.r-project.org/web/packages/OlinkAnalyze/vignettes/bridging_introduction.html#selecting-bridging-samples)
of the Introduction to Bridging tutorial.

## Bridge Sample Selection

## Import NPX files

The input datasets are standard Olink^®^ NPX^TM^ tables. They can be loaded
using `read_NPX()` function with default Olink Software NPX file as
input.

```{r message=FALSE, eval=FALSE, echo = TRUE}
data_E3072 <- read_NPX("~/NPX_Explore3072_location.parquet") # Could also be a CSV file
data_EHT <- read_NPX("~/NPX_ExploreHT_location.parquet")
```

## Checking input datasets and bridging samples

## Normalization

### Is the Assay Bridgeable

### Normalization using the median of paired differences

### Quantile-Quantile Normalization

### Function Output

## Evaluating the quality of bridging

## Export Normalized Data

## FAQs

### Correlation Assays

### Downstream Analysis


## Contact Us

We are always happy to help. Email us with any questions:

-   biostat\@olink.com for statistical services and general stats
    questions

-   support\@olink.com for Olink lab product and technical support

-   info\@olink.com for more information

## Legal Disclaimer

© 2024 Olink Proteomics AB.

Olink products and services are For Research Use Only and not for Use in
Diagnostic Procedures.

All information in this document is subject to change without notice.
This document is not intended to convey any warranties, representations
and/or recommendations of any kind, unless such warranties,
representations and/or recommendations are explicitly stated.

Olink assumes no liability arising from a prospective reader’s actions
based on this document.

OLINK, NPX, PEA, PROXIMITY EXTENSION, INSIGHT and the Olink logotype are
trademarks registered, or pending registration, by Olink Proteomics AB.
All third-party trademarks are the property of their respective owners.

Olink products and assay methods are covered by several patents and
patent applications <https://olink.com/legal/patents>.

