---
title: "Bridging Olink^®^ Explore 3072 to Olink^®^ Explore HT"
output: 
  html_vignette:
    toc: true
    toc_depth: 3
    includes:
      in_header: ../man/figures/logo.html
vignette: >
  %\VignetteIndexEntry{Bridging Olink^®^ Explore 3072 to Olink^®^ Explore HT}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  tidy = FALSE,
  tidy.opts = list(width.cutoff = 95),
  fig.width = 6,
  fig.height = 3,
  message = FALSE,
  warning = FALSE,
  time_it = TRUE,
  fig.align = "center"
)
```

```{r, echo=FALSE}
library(OlinkAnalyze)
library(dplyr)
library(stringr)
library(ggplot2)
library(kableExtra)
```

## Introduction

## Bridge Sample Selection

## Import NPX files

The input datasets are standard Olink^®^ NPX^TM^ tables. They can be loaded
using `read_NPX()` function with default Olink Software NPX file as
input.

```{r message=FALSE, eval=FALSE, echo = TRUE}
data_E3072 <- read_NPX("~/NPX_Explore3072_location.parquet") # Could also be a CSV file
data_EHT <- read_NPX("~/NPX_ExploreHT_location.parquet")
```

## Checking input datasets and bridging samples

## Normalization

### Is the Assay Bridgeable

### Normalization using the median of paired differences

### Quantile-Quantile Normalization

### Function Output

## Evaluating the quality of bridging

## Export Normalized Data

## FAQs

### Correlation Assays

Both the Explore 3072 and Explore HT platforms contain correlation assays in 
their biomarker portfolios. In Explore 3072, these present as overlapping assays 
across panels. In Explore HT, these are overlapping assays across blocks. 
The correlation assays are included for QC purposes, and allow users to evaluate 
data performance across panels in the case of Explore 3072 and across blocks 
in the case of Explore HT. 

Within each platform, the correlation assays contain unique OlinkID values for each 
of their corresponding panels and blocks in Explore 3072 and Explore HT, 
respectively. It should be noted that the correlation assays themselves differ
between each platform. For more information on overlapping assays, you can 
access the [Olink FAQ page](https://olink.com/knowledge/faq). 

Explore 3072 to Explore HT bridging generates and returns adjusted values for all 
data points corresponding to each of the Explore 3072 correlation assays. *For 
bridging purposes, only the Block 5 data points from Explore HT's  correlation assays 
are used as reference values. For downstream analysis, it is recommended to include 
the data from all instances of the correlation assays, barring any clear outliers.* 

### Downstream Analysis


## Contact Us

We are always happy to help. Email us with any questions:

-   biostat\@olink.com for statistical services and general stats
    questions

-   support\@olink.com for Olink lab product and technical support

-   info\@olink.com for more information

## Legal Disclaimer

© 2024 Olink Proteomics AB.

Olink products and services are For Research Use Only and not for Use in
Diagnostic Procedures.

All information in this document is subject to change without notice.
This document is not intended to convey any warranties, representations
and/or recommendations of any kind, unless such warranties,
representations and/or recommendations are explicitly stated.

Olink assumes no liability arising from a prospective reader’s actions
based on this document.

OLINK, NPX, PEA, PROXIMITY EXTENSION, INSIGHT and the Olink logotype are
trademarks registered, or pending registration, by Olink Proteomics AB.
All third-party trademarks are the property of their respective owners.

Olink products and assay methods are covered by several patents and
patent applications <https://olink.com/legal/patents>.

