---
title: "Bridging Olink^®^ Explore 3072 to Olink^®^ Explore HT"
output: 
  html_vignette:
    toc: true
    toc_depth: 3
    includes:
      in_header: ../man/figures/logo.html
vignette: >
  %\VignetteIndexEntry{Bridging Olink^®^ Explore 3072 to Olink^®^ Explore HT}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  tidy = FALSE,
  tidy.opts = list(width.cutoff = 95),
  fig.width = 6,
  fig.height = 3,
  message = FALSE,
  warning = FALSE,
  time_it = TRUE,
  fig.align = "center"
)
```

```{r, echo=FALSE}
library(OlinkAnalyze)
library(dplyr)
library(stringr)
library(ggplot2)
library(kableExtra)
```

## Introduction

## Bridge Sample Selection

## Import NPX files

The input datasets are standard Olink^®^ NPX^TM^ tables. They can be loaded
using `read_NPX()` function with default Olink Software NPX file as
input.

```{r message=FALSE, eval=FALSE, echo = TRUE}
data_E3072 <- read_NPX("~/NPX_Explore3072_location.parquet") # Could also be a CSV file
data_EHT <- read_NPX("~/NPX_ExploreHT_location.parquet")
```

## Checking input datasets and bridging samples

## Normalization

### Is the Assay Bridgeable

### Normalization using the median of paired differences

### Quantile-Quantile Normalization

### Function Output

## Evaluating the quality of bridging

## Export Normalized Data

## FAQs

### Correlation Assays

### Downstream Analysis


## Contact Us

We are always happy to help. Email us with any questions:

-   biostat\@olink.com for statistical services and general stats
    questions

-   support\@olink.com for Olink lab product and technical support

-   info\@olink.com for more information

## Legal Disclaimer

© 2024 Olink Proteomics AB.

Olink products and services are For Research Use Only and not for Use in
Diagnostic Procedures.

All information in this document is subject to change without notice.
This document is not intended to convey any warranties, representations
and/or recommendations of any kind, unless such warranties,
representations and/or recommendations are explicitly stated.

Olink assumes no liability arising from a prospective reader’s actions
based on this document.

OLINK, NPX, PEA, PROXIMITY EXTENSION, INSIGHT and the Olink logotype are
trademarks registered, or pending registration, by Olink Proteomics AB.
All third-party trademarks are the property of their respective owners.

Olink products and assay methods are covered by several patents and
patent applications <https://olink.com/legal/patents>.

