---
title: "Introduction to OlinkAnalyze v5.0"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to OlinkAnalyze v5.0}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Vignettes Have Moved

The detailed vignettes previously hosted in OlinkAnalyze have been moved to a
dedicated package: **OlinkAnalyzeVignettes**. You can install it from CRAN with:

```{r install-vignettes-pkg}
install.packages("OlinkAnalyzeVignettes")
```

Please refer to OlinkAnalyzeVignettes for comprehensive tutorials, advanced
workflows, and detailed examples covering all features of OlinkAnalyze.

This vignette provides a quick-start guide to the key functions introduced in
**OlinkAnalyze v5.0**: reading NPX data, validating and cleaning it with
`check_npx()` and `clean_npx()`, and performing a simple downstream analysis.

## Setup

```{r setup}
library(OlinkAnalyze)
```

## Reading NPX Data

Use `read_npx()` to import data exported from Olink software. The function
supports Excel (`.xlsx`), parquet (`.parquet`), and CSV files.

```{r read-npx}
# Path to the example file shipped with OlinkAnalyze
npx_file <- system.file("extdata",
                         "npx_data_ext.parquet",
                         package = "OlinkAnalyze")

# Read the file into a tibble
npx_df <- read_npx(filename = npx_file)
npx_df
```

For this vignette we will use the built-in `npx_data1` dataset, which is already
in the expected long format:

```{r load-data}
data(npx_data1)
head(npx_data1)
```

## Checking Data Quality with `check_npx()`

`check_npx()` validates the structure and content of your NPX data. It checks
column names, Olink identifiers, missing values, duplicate samples, and QC
warnings, returning a log of potential issues:

```{r check-npx}
check_results <- check_npx(df = npx_data1)
check_results
```

The returned log can be passed directly to `clean_npx()` so that cleaning
decisions are informed by the validation results.

## Cleaning Data with `clean_npx()`

`clean_npx()` removes control samples, flagged assays, QC warnings, and other
problematic entries identified during the check step:

```{r clean-npx}
npx_clean <- clean_npx(
  df = npx_data1,
  check_log = check_results
)
npx_clean
```

You can customise which cleaning steps are applied through the function
arguments (e.g. `remove_qc_warning`, `remove_control_sample`). See
`?clean_npx` for the full list of options.

## Downstream Analysis

With a clean dataset in hand we can move on to statistical testing and
visualisation. The example below uses the `Treatment` column in `npx_data1`,
which has two levels: *Treated* and *Untreated*.

### T-test

`olink_ttest()` performs a Welch two-sample t-test for every protein and
applies Benjamini–Hochberg correction for multiple testing:

```{r ttest}
ttest_results <- olink_ttest(
  df = npx_clean,
  variable = "Treatment",
  check_log = check_results
)
head(ttest_results)
```

### Boxplot

Visualise the top significant proteins with `olink_boxplot()`. First, select a
handful of significant OlinkIDs from the t-test results:

```{r boxplot}
# Select the top significant proteins
top_assays <- ttest_results |>
  dplyr::filter(Threshold == "Significant") |>
  dplyr::slice_head(n = 6) |>
  dplyr::pull(OlinkID)

# Generate the boxplot
olink_boxplot(
  df = npx_clean,
  variable = "Treatment",
  olinkid_list = top_assays,
  number_of_proteins_per_plot = 6
)
```

### Volcano Plot

`olink_volcano_plot()` displays the t-test results as a volcano plot, with
effect size on the x-axis and significance on the y-axis:

```{r volcano}
olink_volcano_plot(p.val_tbl = ttest_results)
```

## Session Info

```{r session-info, eval = TRUE}
sessionInfo()
```
