---
title: "Calculating LOD from Olink® Explore data" 
output: 
  html_vignette:
    toc: true
    toc_depth: 3
    includes:
      in_header: ../man/figures/logo.html
vignette: >
  %\VignetteIndexEntry{Calculating LOD from Olink® Explore data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  tidy = FALSE,
  tidy.opts = list(width.cutoff = 95),
  fig.width = 6,
  fig.height = 3,
  message = FALSE,
  warning = FALSE,
  time_it = TRUE,
  fig.align = "center"
)

library(OlinkAnalyze)
library(dplyr)

```

## Introduction

This tutorial describes how to use Olink® Analyze to integrate Limit of
Detection (LOD) into Olink® Explore HT and Olink® Explore 384/3072
datasets. Although it is recommended to use all Olink Explore data in
downstream analyses, LOD information can be useful when performing
technical evaluations of a dataset.

In this tutorial, you will learn how to use `olink_lod()` to add LOD
information to your Olink Explore dataset. Note that Olink Analyze does
not contain example Olink Explore HT or Olink Explore 384/3072 datasets
within the package, so external data will be necessary for the code
below to work. The external data should contain internal and external
controls for proper calculation and normalization. All file paths should
be replaced with a path to your data and fixed LOD reference file (if
applicable).

## Integrating LOD

Limit of Detection (LOD) is a metric that indicates the lowest
measurable value of a protein. LOD can be helpful when performing
technical evaluations of NPX™ datasets, such as calculating CVs. As a
note, LOD is less important in downstream statistical analyses as values
under LOD typically converge across groups. As such, including data
below LOD is unlikely to increase the risk of false positive
discoveries. Furthermore, data below LOD can be instrumental in
downstream analyses such as biomarker discovery as a protein may be well
expressed in one group and not measured in another group. In this case,
this protein can be a strong biomarker candidate for specific groups.

LOD can be added to Olink Explore NPX datasets using `olink_lod()`. This
function can calculate LOD from an NPX dataset using the dataset's
negative controls or a list of predetermined fixed LOD values (available
in the Document Download Center at [olink.com](https://olink.com/)). As
the default setting, `olink_lod()` will calculate LOD using a dataset's
negative controls.

Olink Explore data is commonly delivered plate control (PC) normalized
or intensity normalized (the normalization type employed is indicated in
the NPX file column Normalization), where the latter is dependent on
that the analyzed samples are randomized. These are reported in the two
respective columns PCNormalizedNPX and NPX. Please notice that for PC
normalized datasets the content in these two columns will be identical,
while for intensity normalized datasets the NPX column will include the
intensity normalized values. Similarly, the `olink_lod()` function adds
two columns to your dataset; PCNormalizedLOD and LOD respectively. For a
PC normalized dataset the content in these two columns will be
identical, while for an intensity normalized dataset the LOD column will
contain intensity normalized LOD values. Examples of results for plate
control and intensity normalization are shown in the tables below.

```{r, echo=FALSE}
set.seed(1234)
table1<-npx_data1 |> 
  head() |> 
  dplyr::select(-c(Index, MissingFreq, Panel_Version, QC_Warning, Subject, Treatment, Site, Time, Project, Panel, PlateID)) |>
  dplyr::mutate(Count = round(NPX * (100+sample(seq(-5,15), size = 1)))) |> 
  dplyr::mutate(SampleType = "SAMPLE") |> 
  dplyr::mutate(Normalization = "Plate control") |> 
  dplyr::mutate(NPX = round(NPX,digits = 2)) |> 
  dplyr::mutate(LOD = round(LOD, digits = 2)) |> 
  dplyr::mutate(PCNormalizedNPX = NPX) |> 
  dplyr::mutate(PCNormalizedLOD = LOD) |> 
  dplyr::select(SampleID, SampleType, OlinkID, UniProt, Assay, Count, NPX,  PCNormalizedNPX,Normalization, LOD, PCNormalizedLOD) 

table1 |> 
  knitr::kable(caption = "Example results from Plate Control Normalized Project") |> 
  kableExtra::kable_styling(font_size = 10)

table1 |>  
  dplyr::mutate(Normalization = "Intensity") |> 
  dplyr::mutate(NPX = round(NPX + 4.16,digits = 2)) |> 
  dplyr::mutate(LOD = round(LOD + 4.16, digits = 2)) |> 
  dplyr::select(SampleID, SampleType, OlinkID, UniProt, Assay, Count, NPX,  PCNormalizedNPX,Normalization, LOD, PCNormalizedLOD) |> 
  knitr::kable(caption = "Example results from Intensity Normalized Project") |> 
  kableExtra::kable_styling(font_size = 10)

```

## Import Olink Explore datasets

Olink Explore datasets are standard Olink Explore HT and Olink Explore
384/3072 NPX tables. The `read_NPX()` function can be used to import an
NPX file in parquet form as generated by Olink® NPX Explore Software.
More information on using `read_NPX()` can be found in [the Olink
Analyze Overview tutorial](Vignett.html).

```{r dataset_generation, eval = FALSE, message=FALSE, warning=FALSE}
explore_npx <- read_NPX("~/Explore_NPX_file.parquet")
```

## Integrating Negative Control LOD

The negative control (NC) LOD method requires at least 10 negative
controls in a dataset. Negative control data is available in the
standard exported Explore HT and Explore 384/3072 NPX parquet files. NCs
can be identified through the SampleID and SampleType columns.

A negative control will not contribute to the minimum number of required
NCs if the negative control does not pass sample QC criteria (sample QC
failure or warning) in all of the data (i.e. all Explore HT blocks, all
Explore 3072 panels, or all Explore 384 panels that were measured)

Negative controls are used to calculate LOD from either PC normalized
NPX or counts. For assays with more than 150 counts in one of the
negative controls, LOD is calculated using the median PC normalized NPX
and adding 3 standard deviations, or 0.2 NPX whichever is larger. For
assays with fewer than 150 counts in all negative controls, LOD is
calculated using the count values which are then converted into PC
normalized NPX.

The resulting LOD is the PC normalized negative control LOD. In the
event that the Explore dataset is intensity normalized, an intensity
normalization adjustment factor is applied and the resulting intensity
normalized LOD is reported in the LOD column and the PC normalized LOD
is reported in the PCNormalizedLOD column.

```{r NCLOD_example, eval = FALSE, message=FALSE, warning=FALSE}
# Integrating negative control LOD for intensity normalized data
explore_npx <- read_NPX("Path_to/Explore_NPX_file.parquet")
olink_lod(explore_npx, lod_method = "NCLOD")
```

## Integrating Fixed LOD

The fixed LOD method uses fixed LOD values that have been calculated on
negative controls used in Olink reference runs using the method
described above for negative control LOD. These values are specific to
the Data Analysis Reference ID, which can be found in your dataset. The
fixed LOD data is available in an external CSV file which can be
downloaded from the Document Download Center at
[olink.com](https://olink.com/). The fixed LOD values reported in this
CSV file are the PC normalized LODs.

The fixed LOD file is read into the `olink_lod()` function to be
integrated into an Explore dataset. In the event that the Explore
dataset is intensity normalized, an intensity normalization adjustment
factor is applied and the resulting intensity normalized LOD is reported
in the LOD column and the PC normalized LOD is reported in the
PCNormalizedLOD column.

```{r FixedLOD, eval = FALSE, message=FALSE, warning=FALSE}
# Reading in Fixed LOD file path into R environment
fixedLOD_filepath <- "Path_to/ExploreHT_fixedLOD.csv"

# Integrating Fixed LOD for intensity normalized data
explore_npx <- read_NPX("~/Explore_NPX_file.parquet")
olink_lod(explore_npx, lod_file_path = fixedLOD_filepath, lod_method = "FixedLOD")
```

## When to use Fixed LOD vs NC LOD

For smaller sized studies (\<10 NCs) we recommend using fixed LOD to
integrate LOD values into your NPX dataset, as LOD calculations on fewer
NCs may provide non-accurate values. However, it is important to keep in
mind that fixed LOD values are not specific to your project, rather
these values are generated by Olink when a new lot of reagents is
released.

For larger projects we recommend calculating LOD from NC to obtain LOD
values that are specific to your project. However, this requires that
the dataset has at least 10 NCs with passing SampleQC.

## Adjusting LOD for Intensity Normalized Data

If an Olink Explore dataset is intensity normalized, a normalization
adjustment factor is applied to the PC normalized LOD within the
`olink_lod()` function.

For each assay, this adjustment factor is calculated as the median NPX
of all samples (excluding Olink's external controls) within each plate.
For Olink Explore 3072, overlapping assays are assessed separately,
within their respective panels. The intensity normalized negative
control LOD is calculated by subtracting this adjustment factor from the
PC normalized negative control LOD.

The intensity normalization LOD adjustment is applied to both the
negative control and fixed LOD methods.

## Export Olink Explore Data with LOD

Olink Explore data with LOD data can be exported using
arrow::write_parquet to export Olink Explore data as a parquet file in
long format.

```{r explore_npx_export, eval = FALSE, message=FALSE, warning=FALSE}
# Exporting Olink Explore data with LOD information as a parquet file
explore_npx <- read_NPX("Path_to/Explore_NPX_file.parquet")

explore_npx_NC_LOD <- explore_npx %>% 
  olink_lod(lod_method = "NCLOD")

# Add metadata for export
df <- explore_npx_NC_LOD |> 
  arrow::as_arrow_table()

df$metadata$FileVersion <- "NA"
df$metadata$ExploreVersion <- "NA"
df$metadata$ProjectName <- "NA"
df$metadata$SampleMatrix <- "NA"
df$metadata$DataFileType <- "Olink Analyze Export File"
df$metadata$ProductType <- "ExploreHT" # "ExploreHT" or "Explore3072"
df$metadata$Product <- "ExploreHT" # "ExploreHT" or "Explore3072"

arrow::write_parquet(x = df, sink = "path_to_output.parquet")
```

## Contact Us

We are always happy to help. Email us with any questions:

-   biostat\@olink.com for statistical services and general stats
    questions

-   support\@olink.com for Olink lab product and technical support

-   info\@olink.com for more information

## Legal Disclaimer

© 2024 Olink Proteomics AB.

Olink products and services are For Research Use Only and not for Use in
Diagnostic Procedures.

All information in this document is subject to change without notice.
This document is not intended to convey any warranties, representations
and/or recommendations of any kind, unless such warranties,
representations and/or recommendations are explicitly stated.

Olink assumes no liability arising from a prospective reader’s actions
based on this document.

OLINK, NPX, PEA, PROXIMITY EXTENSION, INSIGHT and the Olink logotype are
trademarks registered, or pending registration, by Olink Proteomics AB.
All third-party trademarks are the property of their respective owners.

Olink products and assay methods are covered by several patents and
patent applications <https://olink.com/patents/>.
