---
title: "Olink® Analyze Vignette"
author: "Olink DS team"
date: "`r Sys.Date()`"
output: 
  html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Olink® Analyze Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 6,
  fig.height = 3,
  fig.align = "center",
  collapse = TRUE,
  comment = "#>",
  eval = FALSE)

options(tibble.print_min = 4L, tibble.print_max = 4L)
```

> **Note:** The detailed vignettes previously hosted in OlinkAnalyze have been
> moved to a dedicated package: **OlinkAnalyzeVignettes**. You can install it
> from CRAN with `install.packages("OlinkAnalyzeVignettes")`. Please refer to
> OlinkAnalyzeVignettes for comprehensive tutorials, advanced workflows, and
> detailed examples covering all features of OlinkAnalyze.
>
> This vignette presents how to use the key functions in the latest major
> update of **OlinkAnalyze v5.0**.

Olink® Analyze is an R package that provides a versatile toolbox to enable fast and easy handling of Olink® NPX data for your proteomics research. Olink® Analyze provides functions for using Olink data, including functions for importing Olink® NPX datasets, as well as quality control (QC) plot functions and functions for various statistical tests. This package is meant to provide a convenient pipeline for your Olink NPX data analysis.

# Installation

You can install Olink® Analyze from CRAN.

```{r, eval=FALSE}
install.packages("OlinkAnalyze")
```

# Usage

## Load the library

```{r message=FALSE, warning=FALSE}
# Load OlinkAnalyze
library(OlinkAnalyze)

# Load other libraries used in Vignette
library(dplyr)
library(ggplot2)
library(stringr)
```

## Introduction to Olink NPX data format

The package contains two test data files named npx_data1 and npx_data2. These are synthetic datasets that resemble Olink® data accompanied by clinical variables. Olink® data that is delivered in long format or imported with the function *read_NPX* (that converts the data into a long format) contain the following columns:

-   **SampleID** *\<chr\>*: Sample names or IDs.
-   **OlinkID** *\<chr\>*: Unique ID for each assay assigned by Olink. In case the assay is included in more than one panels it will have a different OlinkID in each one.
-   **UniProt** *\<chr\>*: UniProt ID.
-   **Assay** *\<chr\>*: Common gene name for the assay.
-   **MissingFreq** *\<dbl\>*: Missing frequency for the OlinkID, i.e. frequency of samples with NPX value below limit of detection (LOD).
-   **Panel** *\<chr\>*: Olink Panel that samples ran on. Read more about Olink Panels here: <https://olink.com/products/compare>.
-   **Panel_Version** *\<chr\>*: Version of the panel. A new panel version might include some different or improved assays.
-   **PlateID** *\<chr\>*: Name of the plate.
-   **QC_Warning** or **SampleQC** *\<chr\>*: Indication whether the sample passed Olink QC. More information about Olink quality control metrics can be found in our [FAQ](https://olink.com/knowledge/faq?query=quality%20control) by searching "Quality control".
-   **LOD** *\<dbl\>*: Limit of detection (LOD) is the minimum level of an individual protein that can be measured. LOD is defined as 3 times the standard deviation over background.
-   **NPX** *\<dbl\>*: Normalized Protein eXpression, is Olink®'s unit of protein expression level in a log<sub>2</sub> scale. The majority of the functions of this package use NPX values for calculations. Read more about NPX in the Olink [FAQ](https://olink.com/knowledge/faq?query=what%20is%20npx) (Search "What is NPX?") or in Olink's Data normalization and standardization [white paper](https://7074596.fs1.hubspotusercontent-na1.net/hubfs/7074596/05-white%20paper%20for%20website/1096-olink-data-normalization-white-paper.pdf).

**Note:** There are 5 additional variables in the sample datasets npx_data1 and npx_data2 that include clinical or other information, namely: Subject *\<chr\>*, Treatment *\<chr\>*, Site *\<chr\>*, Time *\<chr\>*, Project *\<chr\>*.

The columns found in an Olink data set may vary based on the version and product.

# Reading and preparing NPX data

## Read NPX data (read_NPX)

The read_NPX function imports an NPX file into a tidy format to work with in R. This function supports Olink® NPX files generated by Olink® data software in CSV, Excel, and Parquet formats. No prior alterations to the NPX output file should be made for this function to work as expected.

-   filename: Path to the NPX output file.

```{r message=FALSE, eval=FALSE}
# Read a single NPX file
data <- read_NPX("~/NPX_file_location.xlsx")

# Read the example file shipped with OlinkAnalyze
npx_file <- system.file("extdata",
                         "npx_data_ext.parquet",
                         package = "OlinkAnalyze")
npx_df <- read_NPX(filename = npx_file)
```

For this vignette we use the built-in `npx_data1` dataset, which is already in the expected long format:

```{r message=FALSE, eval=FALSE}
data(npx_data1)
head(npx_data1)
```

## Check NPX data (check_npx)

The `check_npx()` function validates the structure and content of your NPX data. It checks column names, validates Olink identifiers, identifies assays with *NA* values for all samples, and detects duplicate sample identifiers. The returned check log can be passed directly to `clean_npx()` so that cleaning decisions are informed by the validation results.

-   df: NPX data frame in long format.
-   preferred_names: Optional named character vector to assign custom-named columns to internally expected variables. For example, `c("quant" = "PCNormalizedNPX")` to use `PCNormalizedNPX` instead of `NPX`.

```{r message=FALSE, eval=FALSE}
check_results <- check_npx(df = npx_data1)
check_results
```

The function returns a list containing information about column names, invalid OlinkIDs, assays with all *NA* values, duplicate sample IDs, column data types, QC warnings, and UniProt mapping issues.

## Clean NPX data (clean_npx)

The `clean_npx()` function removes control samples, flagged assays, QC warnings, and other problematic entries identified during the check step. It applies a series of cleaning steps including removal of invalid assays, duplicate samples, external control samples, internal control assays, and samples or assays with quality control flags.

-   df: NPX data frame in long format.
-   check_log: Output from `check_npx()`. If not provided, `check_npx()` will be called internally.
-   verbose: Logical. If `TRUE`, prints messages for each cleaning step.

```{r message=FALSE, eval=FALSE}
npx_clean <- clean_npx(
  df = npx_data1,
  check_log = check_results
)
npx_clean
```

You can customise which cleaning steps are applied through the function arguments (e.g. `remove_qc_warning`, `remove_control_sample`, `remove_control_assay`). See `?clean_npx` for the full list of options.

# Downstream analysis

With a clean dataset in hand we can move on to statistical testing and visualisation. The example below uses the `Treatment` column in `npx_data1`, which has two levels: *Treated* and *Untreated*.

## T-test analysis (olink_ttest)

The olink_ttest function performs a Welch 2-sample t-test or paired t-test at confidence level 0.95 for every protein (by OlinkID) for a given grouping variable. It corrects for multiple testing using the Benjamini-Hochberg method ("fdr"). Adjusted p-values are logically evaluated towards adjusted p-value \< 0.05. The resulting t-test table is arranged by ascending p-values.

-   df: NPX data frame in long format should minimally contain protein name (Assay), OlinkID, UniProt, Panel and an outcome factor with 2 levels.
-   variable: Character value that should represent a column in the df to be used as a grouping variable. Needs to have exactly 2 levels.
-   pair_id: Character value indicating which column contains the paired sample identifier. Only used for paired t-tests.

```{r message=FALSE, eval=FALSE}
ttest_results <- olink_ttest(
  df = npx_clean,
  variable = "Treatment",
  check_log = check_results
)
head(ttest_results)
```

## Boxplot (olink_boxplot)

The olink_boxplot function is used to generate boxplots of NPX values stratified on a variable for a given list of proteins.

-   df: NPX data frame in long format should minimally contain protein name (Assay), OlinkID, UniProt and a grouping variable.
-   variable: Single character value indicating the column name to use as a grouping variable in the x axis.
-   olinkid_list: Character vector of OlinkID's that should be used for the boxplot. If not specified, all assays in df are used.
-   number_of_proteins_per_plot: Number of boxplots to include in the facets plot. Default 6.

```{r message=FALSE, eval=FALSE}
# Select the top significant proteins
top_assays <- ttest_results |>
  dplyr::filter(Threshold == "Significant") |>
  dplyr::slice_head(n = 6) |>
  dplyr::pull(OlinkID)

# Generate the boxplot
olink_boxplot(
  df = npx_clean,
  variable = "Treatment",
  olinkid_list = top_assays,
  number_of_proteins_per_plot = 6
)
```

## Volcano plot (olink_volcano_plot)

The olink_volcano_plot function generates a volcano plot using results from the olink_ttest function. The estimated difference is shown in the x-axis and -log<sub>10</sub>(p-value) in the y-axis. The horizontal dotted line indicates p-value = 0.05. Dots are colored based on significance following Benjamini-Hochberg adjustment with a p-value cutoff of 0.05. Significant assays after adjustment can optionally be annotated by OlinkID.

-   p.val_tbl: a data frame of results generated by *olink_ttest*.
-   x_lab: Optional. Character value to use as the x-axis label.
-   olinkid_list: Optional. Character vector of proteins (OlinkID) to label in the plot. If not provided, by default the function will label all significant proteins.

```{r message=FALSE, eval=FALSE}
# select names of top 10 proteins to annotate
top_10_name <- ttest_results |>
  dplyr::slice_head(n = 10) |>
  dplyr::pull(OlinkID)

# volcano plot
olink_volcano_plot(p.val_tbl = ttest_results,
                   x_lab = "Treatment",
                   olinkid_list = top_10_name)
```

## Contact Us

We are always happy to help. Email us with any questions:

-   biostat\@olink.com for statistical services and general stats questions

-   support\@olink.com for Olink lab product and technical support

-   info\@olink.com for more information

## Legal Disclaimer

© 2025 Olink Proteomics AB, part of Thermo Fisher Scientific.

Olink products and services are For Research Use Only. Not for use in diagnostic procedures.

All information in this document is subject to change without notice. This document is not intended to convey any warranties, representations and/or recommendations of any kind, unless such warranties, representations and/or recommendations are explicitly stated.

Olink assumes no liability arising from a prospective reader's actions based on this document.

OLINK, NPX, PEA, PROXIMITY EXTENSION, INSIGHT and the Olink logotype are trademarks registered, or pending registration, by Olink Proteomics AB. All third-party trademarks are the property of their respective owners.

Olink products and assay methods are covered by several patents and patent applications [https://www.olink.com/patents/](https://olink.com/patents/).
