---
title: "Outlier Exclusion Vignette"
author: "Olink DS team"
date: "`r Sys.Date()`"
output: 
  html_vignette:
    toc: true
    toc_depth: 3
    
vignette: >
  %\VignetteIndexEntry{Outlier Exclusion Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

This vignette describes how to use Olink® Analyze to evaluate a dataset for the presence of outliers. When performing statistical analyze it is important to establish the presence of any outlier samples in the data prior to statistical analysis. There are many reasons why a sample might be an outlier, it could be due to a data entry or measurement error (i.e. labeling a control sample as a disease sample), sampling problems or unusual conditions (i.e. contamination), or natural variation. In many parametric statistics tests the mean of each group is used to determine differences between groups. Since the mean is highly sensitive to outliers, it is important to examine datasets for outliers prior to analysis as these outliers could have a large influence on the statistical results.

In Olink Analyze, there are three visualization functions that can be used to identify potential outlier samples. In this vignette, you will learn how to use `olink_pca_plot()`, `olink_dist_plot()` , and `olink_qc_plot()` to identify and remove outliers from a dataset.

```{r Outlier_data, include=FALSE}
library(OlinkAnalyze)
knitr::opts_chunk$set(fig.height=5,fig.width=6)
knitr::opts_chunk$set(warning = F)
knitr::opts_chunk$set(message = F)
knitr::opts_chunk$set(fig.pos = 'H')
knitr::opts_chunk$set(fig.align="center")
knitr::opts_knit$set(eval.after = "fig.cap")
# Create Dataset with outliers
outlier_data <- npx_data1 |> 
  dplyr::mutate(NPX = ifelse(SampleID == "A25", NPX + 4, NPX)) |> 
  dplyr::mutate(NPX = ifelse(SampleID == "A52", NPX - 4, NPX)) |> 
  dplyr::filter(!stringr::str_detect(SampleID, "CONTROL"))

group_data <- npx_data1 |> 
  dplyr::mutate(NPX = ifelse(Site == "Site_D", NPX + 3, NPX)) |> 
  dplyr::filter(!stringr::str_detect(SampleID, "CONTROL"))

 
```

## PCA Plot

### Overview of PCA Plot

Principal Component Analysis (PCA) is a dimensional reduction technique. PCA plots can been particularly useful to visualize variability within high dimensional data by plotting the data along the axes of greatest variation. The first two principal components make up the largest axes of variance between samples. In `olink_pca_plot()` samples are clustered together based on similarities in overall expression patterns of the measured proteins. These samples can be colored by any categorical variable to determine which biological factors may be contributing to global effects across the samples. PCA plots can be used to:

-   Identify individual outlier samples (**Figure 1A**)

-   Identify groups of outliers (**Figure 1B**)

-   Identify batch effects (see Bridging Vignette)

Regardless of what the PCA is being used for, PCA plots are great tools to get an overview of the data prior to analysis and pick up on any global trends.

```{r Outlier_Example, fig.cap=fcap}
p1<- outlier_data |> olink_pca_plot(label_samples = T, quiet = T)
p2<- group_data |> olink_pca_plot(color_g = "Site", quiet = T)
ggpubr::ggarrange(p1[[1]], p2[[1]], nrow = 2, labels = "AUTO")
fcap <- "**Figure 1** **A.** PCA can be used to identify individual outlier samples as shown by samples A25 and A52. **B.** PCA can be used to identify difference in groups as seen by Site_D samples."
```

 ### How to use PCA plot to identify an outlier

As shown above, PCA plots can be used to identify global differences in specific samples or sets of samples. Often times these samples can be attributed to natural variations, but sometimes these samples are outliers due to technical or sample specific issues. To determine if a sample is a true outlier and should be excluded, consider the following:

-   Did the sample pass QC? - Sometimes samples that are outliers and have a QC warnings may indicate sample or technical issues. For example, a buffer sample will often flag on QC and be plotted as an outlier as there are no proteins in the sample.

-   How far is the sample from other samples? - Samples variability within a specific group may be larger or there may be global variables within a group. In this case it is important to consider the sample within the context of the project.

-   Does the sample appear as an outlier by other plots? - If a sample is an outlier in the PCA, NPX distribution, and QC plots, that this might indicate a true outlier.

-   Is the sample an outlier on all panels? - Some samples may perform better on specific samples. In the next section we will explain how to view samples by panel.

PCA plots can be generated using `olink_pca_plot()` and specifying the color for each sample using the `color_g` argument. By default the samples will be colored by QC_Warning. Prior to generating the PCA plot, the duplicate SampleIDs (often Control samples) will need to be renamed or filtered out. 

```{r}
OlinkAnalyze::npx_data1 |> 
  dplyr::filter(stringr::str_detect(SampleID, "CONTROL", negate = T)) |> # Filter duplicate SampleIDs
  olink_pca_plot(color_g = "Treatment")
```
In this dataset there do not appear to be any outliers in the PCA plot.

### PCA plot by panel 

When multiple panels are run, there is a chance a sample may be an outlier on one panel. To get a global view of the samples per Panel, the `byPanel` argument can be specified. 

```{r}
OlinkAnalyze::npx_data2 |> 
  dplyr::filter(stringr::str_detect(SampleID, "CONTROL", negate = T)) |> # Filter out control SampleIDs
  olink_pca_plot(byPanel = TRUE) # Specify by panel
```

The PCA plots will be saved in a list of ggplot objects and each can be viewed individually using ``[[n]]`` syntax as shown below.

```{r}
pca_plots<-OlinkAnalyze::npx_data2|> # Save the PCA plot to a variable
  dplyr::filter(stringr::str_detect(SampleID, "CONTROL", negate = T)) |> # Filter duplicate SampleIDs
  olink_pca_plot(byPanel = TRUE, quiet = TRUE) # By panel, do not print ggarranged plot 
pca_plots[[1]] #Cardiometabolic PCA
pca_plots[[2]] #Inflammation PCA
```

The plots above show an example where there are not any clear outliers in the PCA. For the purposes of this vignette, we have also generated data where 2 samples appear as outliers.

```{r}
outlier_data |> 
  dplyr::filter(stringr::str_detect(SampleID, "CONTROL", negate = T)) |> # Filter duplicate SampleIDs
  olink_pca_plot(byPanel = TRUE) 
```

However from this plot, we can not identify which to samples are outliers. In the next section we will go over how to label outliers in the plot.


### Labeling outliers

There are two ways to label the samples in the PCA plot. The first way is to use the `label_samples` argument. This argument will label all samples by replacing the dot with the SampleID. 

```{r}
outlier_data |> 
  dplyr::filter(stringr::str_detect(SampleID, "CONTROL", negate = T)) |> # Filter duplicate SampleIDs
  olink_pca_plot(label_samples = TRUE) 
```
Here we can see that samples A25 and A52 appear as outliers. This method can be useful when there is clear separation in samples. However it can be difficult to identify additional trends and all samples are labeled. In this case the samples much also be visually identified as opposed to programmatically extracted from the plot. For a cleaner plot with only the outliers labeled, we can use `outlierDefX`, `outlierDefY`, `outlierLines`, and `label_outliers`. These arguments will plot lines at a number of standard deviations from the mean of the plotted PC and label the samples outside of these lines. The values of `outlierDefX` and `outlierDefY` will need to be generated by the users and may require some manual adjustment to determine the correct number to highlight the outliers.

```{r}
outlier_data |> 
  dplyr::filter(stringr::str_detect(SampleID, "CONTROL", negate = T)) |> # Filter duplicate SampleIDs
  olink_pca_plot(outlierDefX = 3, outlierDefY = 3, 
                 outlierLines = TRUE, label_outliers = TRUE) 
```


To remove the lines and just keep the outliers labelled `outlierLines` can be set to False.

```{r}
outlier_data |> 
  dplyr::filter(stringr::str_detect(SampleID, "CONTROL", negate = T)) |> # Filter duplicate SampleIDs
  olink_pca_plot(outlierDefX = 3, outlierDefY = 3, 
                 outlierLines = FALSE, label_outliers = TRUE) 
```

Once the correct outliers have been highlighted in the graph, we can then programmatically extract the outlier SampleIDs using *dplyr*.

```{r}
outliers_pca_labeled <- outlier_data |> 
  dplyr::filter(stringr::str_detect(SampleID, "CONTROL", negate = T)) |> # Filter duplicate SampleIDs
  olink_pca_plot(outlierDefX = 3, outlierDefY = 3, outlierLines = FALSE, 
                 label_outliers = TRUE, quiet = TRUE) 

outliers_pca_labeled[[1]]$data |> 
  dplyr::filter(Outlier == TRUE) |> 
  dplyr::select(SampleID) |> 
  dplyr::distinct()

```


## NPX Distribution Plot

### Overview of NPX Distribution Plot

### How to use NPX distribution plot to identify an outlier

### NPX Distribution plot by panel

## QC Plot (IQR vs Sample Median)

### Overview of QC Plot

### How to use QC plot to identify an outlier

### QC plot by panel

### Labeling outliers with lines

### Extracting outliers names from data

## Interpreting and Handling Outliers

### When should outliers be removed

## Contact Us!

We are always happy to help. Email us with any questions:

biostat\@olink.com for statistical services and general stats questions

biostattools\@olink.com for Olink Analyze and Shiny app support

support\@olink.com for Olink lab product and technical support

info\@olink.com for more information
