---
title: "Olink^®^ Analyze Plate Randomization Vignette"
author: "Olink Data Science"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: false # replaced by \tableofcontents
    # toc_depth: 3
    # number_sections: true
    fig_caption: true
    latex_engine: xelatex
header-includes:
  - \usepackage{graphicx}
  - \usepackage{float}
  - \usepackage{fontspec} 
  - \setmainfont{Swedish Gothic}  
  - \usepackage{setspace}
  - \onehalfspacing    # for 1.5x line spacing, options \singlespacing, ...
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhead{}
  - \fancyhead[L]{\includegraphics[width=20pt]{../man/figures/small_logo.png}}
  - \fancyhead[C]{\textbf{Plate Randomization Vignette}}
  - \fancyhead[R]{\thepage}
  - \setlength{\headheight}{23.74pt}
  - \addtolength{\topmargin}{-11.7pt}
  - \renewcommand{\headrulewidth}{0.4pt}
  - \fancyfoot{}  # clears footer
vignette: >
  %\VignetteIndexEntry{Plate Randomization Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- adding logo after title, author and date, and before toc -->

\thispagestyle{empty}
\begin{center}
\vspace*{2cm}
\includegraphics[width=2cm]{../man/figures/Olink\-Symbol_Black.png}
\end{center}
\newpage
\pagestyle{fancy}
\setcounter{tocdepth}{2}
\tableofcontents
\clearpage

```{r, echo=FALSE,message=FALSE, eval=FALSE}
library(OlinkAnalyze)
```

```{r include=FALSE}
library(OlinkAnalyze)
data(manifest)
```


```{r, echo=FALSE,message=FALSE}
library(dplyr)
library(ggplot2)
```

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE,
                      fig.height=3,
                      fig.width=6,
                      message = FALSE) 
```

## Introduction

This tutorial describes how to use Olink® Analyze to randomize samples across plates with an option to keep samples from the same subject on the same plate. When a study is well randomized, the experimental variables can be considered to be evenly distributed across each plate, even for a larger study. Proper sample randomization will empower your study and minimize the risk of introducing any bias that can confound downstream analyses. Samples from the same subject, for example, those that are taken during a longitudinal research should all be placed on the same plate to further limit data fluctuation and preserve the paired nature of the samples. In order to evenly disperse the remaining experimental variables between plates, the individuals are then distributed across plates. If randomization is not performed, true biological variation may be missed or misidentified, for example the true biological variation could be misidentified as technical variations. In most cases, a complete randomization will be used, as it is difficult to control for every variable without biasing the randomization.To check the quality of the randomization, outcomes can be assessed by visualizing the layout of the plates. In this tutorial, you will learn how to achieve this using the `olink_plate_randomizer()` function. And on top of that, you will also learn how to use `olink_displayPlateLayout()` and `olink_displayPlateDistributions()` to evaluate the performed randomization based on a given grouping variable.


## Sample randomization

The input manifest should be a tibble/data frame in long format containing all sample ID's. Sample ID column must be named SampleID. An example manifest is shown in Table 1. 

```{r manifest, echo = F, message=FALSE}
 manifest %>% head(10)  %>% 
  knitr::kable(
               booktabs = T,
               linesep = "",
               digits = 4,
               longtable = T,
               row.names = F,
               caption = "Example of Sample manifest",
               label = "manifest")
```


### Complete randomization

The simplest way to randomize the samples is to perform complete randomization.


```{r complete_randomization, message=FALSE, warning=FALSE,results='hide'}
randomized.manifest <- olink_plate_randomizer(manifest, seed=123456)
```


### Generate randomization scheme that keeps subjects on the same plate

In the case of multiple samples per subject (e.g. in longitudinal studies), Olink recommends keeping each subject on the same plate to further reduce variation in the data. The individuals should then be distributed so that the remaining experimental variables of interest are evenly distributed across plates. This could be achieved by using the `SubjectColumn` argument. The `SubjectColumn` argument must correspond to a column within the manifest and cannot be SampleID. If `SubjectColumn` is specified, every SampleID must have a value in the subject column, even if there is only 1 sample for that subject. If there are too many samples per subject (>8), complete randomization is recommended. 


The number of samples on each plate can be specified via the `available.spots` argument. For example, the following code will lead to 78 samples on the first plate, and 30 samples on the second and the third plates. The number of iterations for fitting subjects on the same plate can be set by the `iterations` argument. 


```{r subjects_randomization, message=FALSE,results='hide'}
randomized.manifest <- olink_plate_randomizer(manifest,
                                                SubjectColumn="SubjectID", 
                                                available.spots=c(48,48,42), 
                                                iterations = 500,
                                                seed=123456)
```

The plate size can be either directly set to 48 or 96 using the `PlateSize` argument or can be inferred by setting the `Product` argument to "Target 96", "Target 48", "Explore 384","Explore 3072", "Reveal", "Flex", or "Explore HT." The number of Olink external controls can be set using the `num_ctrl` argument, which defaults to 8 controls. The Olink external controls can also be set to be randomized throughout the plate using the `rand_ctrl` argument.

```{r randomize_controls, message = FALSE, eval = FALSE}
randomized_manifest <- olink_plate_randomizer(manifest, 
                                              Product = "Explore HT",
                                              SubjectColumn = "SampleID",
                                              num_ctrl = 10, 
                                              rand_ctrl = TRUE)
```

```{r, rand_ctrl_code, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
randomized_manifest <- olink_plate_randomizer(manifest, 
                                              Product = "Explore HT",
                                              SubjectColumn = "SampleID",
                                              num_ctrl = 10, 
                                              rand_ctrl = TRUE)
```

```{r randomized_controls_table, echo=FALSE, message=FALSE, warning=FALSE}
randomized_manifest %>% 
  head(10)  %>% 
  knitr::kable(
               booktabs = T,
               linesep = "",
               digits = 4,
               longtable = T,
               row.names = F,
               caption = "Example of Randomized Sample manifest",
               label = "manifest")

```

## Multi-study manifests

In the case of a manifest that includes multiple studies, the optional `study` argument can be used to group samples by study and randomize the samples within the study. If a column has the name `study`, within study randomization will be performed as well. In the example below, "Site" is used as a surrogate for study. After randomization the sites are kept together across the plates.

```{r}
randomized.manifest_c <- olink_plate_randomizer(manifest, study = "Site")
olink_displayPlateLayout(randomized.manifest_c, fill.color = "Site")
```


## Visualization

`olink_displayPlateLayout()` and `olink_displayPlateDistributions` functions can be used to assess the quality of the randomization.

### Plate layouts

The `olink_displayPlateLayout()` function displays the layouts of the plate and based on the variable of interest, which can be specified using the `color_g` argument. The `include.label` argument can be set to `TRUE` to label the variable of interest on the plot directly.

```{r figure_captions, message=FALSE, echo=FALSE}
fcap1 <- "Randomized samples in 96 well plate format, colored by Subject ID."
fcap2 <- "Randomized samples in 96 well plate format with labeled wells."
fcap3 <- "Distribution of Subject ID across randomized plates."
fcap4 <- "Distribution of Site across randomized plates."

```



```{r, out.width = "\\linewidth", fig.height = 6, fig.width = 7.5, fig.align = "center", fig.cap= fcap1}
olink_displayPlateLayout(randomized.manifest, fill.color = 'SubjectID', include.label = FALSE)
```


```{r, out.width = "\\linewidth", fig.height = 6, fig.width = 7.5, fig.align = "center", fig.cap= fcap2}
olink_displayPlateLayout(randomized.manifest, fill.color = 'SubjectID', include.label = TRUE)
```

### Plate distribution

The distribution of the given grouping variable on each plate can also be visualized using the function \linebreak `olink_displayPlateDistributions()` to generate a bar chart. For example, by setting `fill.color = 'SubjectID'`, we can confirm all the samples from the same subject were put on the same plate. 

```{r, fig.align = "center", fig.cap=fcap3}
olink_displayPlateDistributions(randomized.manifest, fill.color = 'SubjectID')
```

Additionally, this function can be used to visualize the distribution of other variables. For examples, the distribution of the variable `Site` can be visualized by setting `fill.color = 'Site'`. 

```{r, fig.align = "center", fig.cap= fcap4, fig.pos="H"}
olink_displayPlateDistributions(randomized.manifest, fill.color = 'Site')
```

If the visualizations indicate the samples have been poorly randomized, the randomization can be performed again with different value for `seed` and/or `iterations`. 

## Data Output

The output of the randomization function is a "tibble" including SampleID and the corresponding values from the other columns in the input manifest to well positions. Columns include same columns as Manifest with additional columns:

  - plate: Plate number

  - column: Column on the plate

  - row: Row on the plate

  - well: Well location on the plate

The randomized manifest can be exported as any file type based on the downstream need. For example, the file can be exported as an Excel file using the `write_xlsx` function from Package `writexl`.

```{r, eval = FALSE}
library(writexl)
write_xlsx(randomized.manifest,"randomized.manifest.xlsx")
```

\clearpage

## Contact Us

We are always happy to help. Email us with any questions:

-   biostat\@olink.com for statistical services and general stats
    questions

-   support\@olink.com for Olink lab product and technical support

-   info\@olink.com for more information

## Legal Disclaimer

© 2025 Olink Proteomics AB, part of Thermo Fisher Scientific.

Olink products and services are For Research Use Only. Not for use in diagnostic procedures.

All information in this document is subject to change without notice. This document is not intended to convey any warranties, representations and/or recommendations of any kind, unless such warranties, representations and/or recommendations are explicitly stated.

Olink assumes no liability arising from a prospective reader’s actions based on this document.

OLINK, NPX, PEA, PROXIMITY EXTENSION, INSIGHT and the Olink logotype are trademarks registered, or pending registration, by Olink Proteomics AB. All third-party trademarks are the property of their respective owners.

Olink products and assay methods are covered by several patents and patent applications [https://www.olink.com/patents/](https://olink.com/patents/).
